Space-time plots

### 2017-05-24 Rmd/space-time.Rmd
Updated analysis of annual change in observing methods and scales. Revision of main text to accomodate this change, as well as addition of related figures to SI.  Fixes on explanation of methods/FAQ in SI.

# Space versus time analysis
```{r, message = FALSE, warning=FALSE, results='hold'}
library(ecoscales)
library(readxl)
# library(stringdist)
library(maptools)
library(RColorBrewer)
library(viridis)

p_root <- set_base_path("ecoscales")
p_dat <- full_path(p_root, "external/data/result")
p_calib <- full_path(p_root, "external/data/calibration")
# dat <- as.data.table(read_excel(full_path(p_dat, "merged.xlsx")))
dat <- as.data.table(read_excel(full_path(p_dat, "mergedffffff.xlsx")))
# dato <- as.data.table(read_excel(full_path(p_dat, "archive/mergedfff.xlsx")))
cal <- data.table(read_excel(fp(p_calib, "merged_calibrationffffff.xlsx")))
full <- data.table(read_excel(fp(p_dat, "full_no_treuer_toss.xlsx")))
# full <- data.table(read_excel(fp(p_dat, "full_less_Treuer.xlsx")))

@@

## Data prep
### Merge full and calibration datasets
```{r, message=FALSE, warning=FALSE, results='hide'}
# dat2 <- copy(dat[, .(st, DOI, study_type, plot_res, n_sites, act_ext, eff_ext,
#                      samp_dur, t_btwn_samp, act_dur, eff_dur)])
# rnms <- c("st", "DOI", "study_type", "feature", "observer", "plot_res", 
#           "n_sites", "act_ext", "eff_ext", "samp_dur", "t_btwn_samp",
#           "act_dur", "eff_dur")
dat2 <- copy(dat[, .(DOI, study_type, plot_res, n_sites, act_ext, eff_ext,
                     samp_dur, t_btwn_samp, act_dur, eff_dur)])
selind <- 3:ncol(dat2)
for(j in names(dat2)[selind]) set(dat2, j=j, value = as.numeric(dat2[[j]]))
rnms <- c("DOI", "study_type", "feature", "observer", "plot_res", 
          "n_sites", "act_ext", "eff_ext", "samp_dur", "t_btwn_samp",
          "act_dur", "eff_dur")
calr <- cal[, rnms, with = FALSE]
selind <- c(3, 5:length(rnms))
for(j in names(calr)[selind]) set(calr, j=j, value = as.numeric(calr[[j]]))

@@

>>>>

# Space versus time analysis
```{r, message = FALSE, warning=FALSE, results='hold'}
library(ecoscales)
library(readxl)
# library(stringdist)
library(maptools)
library(RColorBrewer)
library(viridis)
# library(zyp)

p_root <- set_base_path("ecoscales")
p_dat <- fp(p_root, "external/data/result")
p_calib <- fp(p_root, "external/data/calibration")
p_fig <- fp(p_root, "paper/figures")

# dat <- as.data.table(read_excel(full_path(p_dat, "merged.xlsx")))
dat <- as.data.table(read_excel(full_path(p_dat, "mergedffffff.xlsx")))
# dato <- as.data.table(read_excel(full_path(p_dat, "archive/mergedfff.xlsx")))
cal <- data.table(read_excel(fp(p_calib, "merged_calibrationffffff.xlsx")))
full <- data.table(read_excel(fp(p_dat, "full_no_treuer_toss.xlsx")))
# full <- data.table(read_excel(fp(p_dat, "full_less_Treuer.xlsx")))

@@

## Data prep
### Merge full and calibration datasets
```{r, message=FALSE, warning=FALSE, results='hide'}
# dat2 <- copy(dat[, .(st, DOI, study_type, plot_res, n_sites, act_ext, eff_ext,
#                      samp_dur, t_btwn_samp, act_dur, eff_dur)])
# rnms <- c("st", "DOI", "study_type", "feature", "observer", "plot_res", 
#           "n_sites", "act_ext", "eff_ext", "samp_dur", "t_btwn_samp",
#           "act_dur", "eff_dur")
dat2 <- copy(dat[, .(DOI, study_type, plot_res, n_sites, act_ext, eff_ext,
                     samp_dur, t_btwn_samp, act_dur, eff_dur)])
# selind <- 3:ncol(dat2)
selind <- 4:ncol(dat2)
for(j in names(dat2)[selind]) set(dat2, j=j, value = as.numeric(dat2[[j]]))
rnms <- c("DOI", "study_type", "feature", "observer", "plot_res",
          "n_sites", "act_ext", "eff_ext", "samp_dur", "t_btwn_samp",
          "act_dur", "eff_dur")
calr <- cal[, rnms, with = FALSE]
selind <- c(3, 5:length(rnms))
for(j in names(calr)[selind]) set(calr, j=j, value = as.numeric(calr[[j]]))

@@

### Study type by year

Calculate types of study by year, as total and percent
```{r, warning = FALSE, message = FALSE, fig.height=3}
# One Oecologia study listed as 2004 (by Endnote) was actually 2003 
# Set it to 2004
stypes <- copy(dat)  # new dataset out of dat (which retain year info)
stypes[study_year < 2004, study_year := 2004]

# bring in study_type from calr set, grabbing year from original cal dt
calyr <- cal[DOI %in% calr[, unique(DOI)], unique(study_year), by = DOI]
calyr[, V1 := as.numeric(V1)]
calyr <- calyr[, list("study_year" = mean(V1, na.rm = TRUE)), by = DOI]
calyr <- merge(calyr, calr[, list(DOI, study_type)], by = "DOI")

# bind
stypes <- rbind(stypes[, list(DOI, study_year, study_type)], calyr)
na_effext <- datf[is.na(eff_ext), unique(DOI)]  # ID NA eff_ext values

# Estimate and plot study type by year
stypeyr <- stypes[, list(.N, study_type), by = study_year][, {
  list("tot" = N, "ct" = .N, "prop" = .N / N * 100)
}, by = .(study_type, study_year)][order(study_year)]
stypeyr <- stypeyr[, lapply(.SD, mean), .SDcols = c("tot", "ct", "prop"), 
                   by = .(study_type, study_year)]
```
```{r, eval = FALSE}
obs_types <- c("remote", "field", "auto")
obs_types2 <- c("Remote sensing", "Field", "Automatic")
cols <- c("orange", "green4", "blue4")
png(fp(p_fig, "stype_by_yr.png"), width = 7, height = 2.75, res = 300, 
    units = "in")
par(mfrow = c(1, 3), oma = c(0, 3, 0, 0), mar = c(4, 1, 2, 1))
for(i in 1:3) {  # i <- 1
  plot(c(2004, 2014), c(0, 100), xlab = "", ylab = "", pch = "")
  stypeyr[like(study_type, obs_types[i]), {
    points(study_year, prop, col = cols[i], pch = 20)
    yr = as.numeric(study_year)
    slm = lm(prop ~ yr, weights = tot)  # weighted regression
    slms = summary(slm)
    abline(slm, col = cols[i])
    cf = round(coef(slm)[1:2], 4)
    pf = round(coef(slms)[8], 2)
    rf = round(slms$adj.r.squared, 2)
    cfa = paste0(cf[2], "% year")
    text(x = 2006, y = 100, labels = substitute(x^-1, list(x = cfa)), 
         cex = 0.9)
    text(x = 2011.5, y = 100, cex = 0.9,
         labels = substitute(R^2~"="~x~"; p <"~y, list(x = rf, y = pf)))
    if(i == 1) mtext("percent", side = 2, line = 2.5, cex = 0.8)
    if(i == 2) mtext("year", side = 1, line = 2.5, cex = 0.8)
    mtext(obs_types2[i], side = 3, cex = 0.8)
  }]
}
dev.off()
```

Extrapolating percentage of studies using remote sensing to 2017
```{r}
preds <- stypeyr[like(study_type, "remote"), {
  slm = lm(prop ~ study_year, weights = tot)
  predict(slm, newdata = data.frame("study_year" = 2004:2017))
}]
names(preds) <- 2004:2017
```

Percent of remote sensing studies by year
```{r}
round(preds, 1)
```

Average annual percent of studies using remote sensing 2004-2014, and projected increases in this percent through 2017 (2004-2017) 
```{r}
# murs2014 <- stypeyr[study_type == "remote", round(mean(prop), 2)]
murs2014 <- stypeyr[, sum(ct[study_type == "remote"]) / sum(ct) * 100]
pctinc <- mean(preds) / mean(preds[1:(length(preds) - 3)])

pcttab <- cbind("2014" = murs2014, "pct increase" = (pctinc - 1) * 100,
                "2017" = murs2014 * pctinc)
round(pcttab, 1)
```

```{r}
types <- datf[, .N, by = study_type]
rs2017 <- mean(preds)  # avg predicted % using RS through 2004-2017
rs2014 <- mean(preds[1:(length(preds) - 3)])  # avg predicted % RS 2004-2014
# rs2017 / rs2014

# translate to N more RS studies would have been selected given those %ages
rsinc <- rs2017 / rs2014 * types[study_type == "remote", N] - 
                 types[study_type == "remote", N]

remote <- datf[study_type == "remote", eff_ext]
notremote <- datf[study_type != "remote", eff_ext]
# a <- (length(remote) + rsinc) / (nrow(datf))
# b <- length(remote) / nrow(datf)
# a / b
# ((length(remote) + rsinc) / (length(notremote)) / length(b))

# (rsext * rs2017 + nonrsext * (1 - rs2017)) / 
#   (rsext * rs2014 + nonrsext * (1 - rs2014))

set.seed(1)
ext2017 <- unlist(lapply(1:1000, function(x) {  # x = 1
  ind <- sample(1:length(remote), size = round(rsinc))
  mean(c(remote, remote[ind], notremote), na.rm = TRUE)
}))

dec1417 <- log10(mean(ext2017, na.rm = TRUE) / 
                   datf[, mean(eff_ext, na.rm = TRUE)])
qtiles1417 <- log10(quantile(ext2017, c(0.025, 0.975), na.rm = TRUE) / 
                      datf[, mean(eff_ext, na.rm = TRUE)])

pctdiff1417 <- mean(ext2017, na.rm = TRUE) / datf[, mean(eff_ext, na.rm = TRUE)]
pctqtiles1417 <- quantile(ext2017, c(0.025, 0.975), na.rm = TRUE) / 
                      datf[, mean(eff_ext, na.rm = TRUE)]

cbind("decade" = round(dec1417, 4), "dec2.5" = round(qtiles1417[1], 4),
      "dec97.5" = round(qtiles1417[2], 4),
      "pctdiff" = round((pctdiff1417 - 1) * 100, 1),
      "pctdiff2.5" = round((pctqtiles1417[1] - 1) * 100, 1), 
      "pctdiff97.5" = round((pctqtiles1417[2] - 1) * 100, 1))
```

### Change in dimensions by year
```{r}
# Examines and plot trends in dimensions
doiyr <- rbind(unique(dat[, .(DOI, study_year)]), 
               unique(cal[, .(DOI, study_year)]))
doiyr <- doiyr[study_year %in% 2004:2014]
datf2 <- copy(datf)
datf2 <- merge(doiyr, datf2, by = "DOI")  # collect year of publication
# datf2[, N := .N, by = study_year]
```
```{r, eval = FALSE}
vars <- c("plot_res", "eff_ext", "t_btwn_samp", "eff_dur")
varlabs <- c("Resolution", "Extent", "Interval", "Duration")
labcols <- c(brewer.pal(8, "Reds")[c(7, 7)], brewer.pal(8, "Blues")[c(7, 7)])
png(fp(p_fig, "dim_by_yr.png"), width = 7, height = 6, res = 300, 
    units = "in")
par(mfrow = c(2, 2), mar = c(1, 3, 1, 1), oma = c(2, 3, 0, 0))
for(i in 1:length(varlabs)) {  # i <- 1
  if(i == 3) {
    DT <- copy(datf2[t_btwn_samp != 0, ])
  } else {
    DT <- copy(datf2)
  }
  DT[, list(mean(get(vars[i]), na.rm = TRUE), .N), by = study_year][, {
    dim = log10(V1)
    yr = as.numeric(study_year)
    if(i %in% 1:2) rng = c(0, 10)
    if(i %in% 3:4) rng = c(0, 5)
    if(i %in% 3:4) {
      xlabs = seq(2004, 2014, 2)
    } else {
      xlabs = rep("", 6)
    }
    plot(yr, dim, ylim = rng, xaxt = "n", pch = 20, col = labcols[i], 
         main = varlabs[i], mgp = c(2, 0.5, 0), tcl = -0.2, ylab = "")
    axis(1, at = seq(2004, 2014, 2), labels = xlabs, tcl = -0.2, 
         mgp = c(2, 0.5, 0))
    slm = lm(dim ~ yr, weights = N)
    slms = summary(slm)
    abline(slm, col = labcols[i])
    cf = round(coef(slm)[1:2], 3)
    pf = round(coef(slms)[8], 2)
    rf = round(slms$adj.r.squared, 2)
    text(x = 2006.25, y = rng[2] * 0.97, cex = 0.9,
         labels = substitute(x~year^-1, list(x = cf[2])))
    text(x = 2011.5, y = rng[2] * 0.97, cex = 0.9,
         labels = substitute(R^2~"="~x~"; p <"~y, list(x = rf, y = pf)))
    if(i == 1) mtext(expression("log10 m"^2), side = 2, line = 2)
    if(i == 2) mtext("log10 ha", side = 2, line = 2)
    if(i %in% 3:4) mtext("log10 days", side = 2, line = 2)
  }]
}
dev.off()
```

Percent increase in average extent by year
```{r}
preds <- datf2[, list(mean(eff_ext, na.rm = TRUE), .N), by = study_year][, {
  dim = log10(V1)
  yr = as.numeric(study_year)
  slm = lm(dim ~ yr, weights = N)
  predict(slm, newdata = data.frame("yr" = 2004:2017))
}]
# preds[14] - preds[13]

# projected percent increase in extent 2014-2017
pctinc <- mean(preds) / mean(preds[1:(length(preds) - 3)])
muext2014 <- datf2[, mean(eff_ext, na.rm = TRUE)]
pcttab <- cbind("2014" = log10(muext2014), "pct increase" = (pctinc - 1) * 100,
                "2017" = log10(muext2014 * pctinc), 
                "decade" = log10((pctinc * muext2014) / muext2014))
round(pcttab, 3)
```
